---
title: "A Practice of General Form of Luria-Delbrück Distribution"
output:
  pdf_document: default
  html_notebook: default
---
# System Dynamics in Biology Workshop: Exercise Day 2

## Speaker: Professor Arthur Lander

*Course material and exercise are originally in Wolfram Language. This is a practice to get some hands-on experience in R.*

> You are working on a line of tumor cells in cell culture. At equilibrium, **95%** of the cells express EGFR and **5%** do not. If the cells are switching back and forth between two stable states, the model could be formulated as:
$$[EGFR^+] \underset{q}{\overset{p}{\rightleftharpoons}} [EGFR^-]$$
Assuming the culture is in equilibrium, what can we say about the ratio of p and q?

Let the culture in equilibrium, from the laws of mass action, we could assume:

$$\frac{d[EGFR^+]}{dt} = p[EGFR^+] = 0 = q [EGFR^-] = \frac{d[EGFR^-]}{dt}$$

and as a result:

$$\frac{p}{q} = \frac{[EGFR^-]}{[EGFR^+]} = \frac{5}{95}$$

> In order to find out the values of *p* and *q*, I could have set up a complicated time-lapse experiment. On the other hand, your friend suggest that you could set up a culture starting from single cell for 12 dividing times, repeat for 500 times, and check the variance-mean ratio of EGFR^+^ cell numbers. The friend suggests that this would also give the value of *p* and *q*.

The question could be further boiled down to: **Given that we know the relationship between p and q, whether a specific p corresponds to a specific variance-mean ratio in this experiment?**

To understand that, we set up a simulation of this experiment, and see if there's such a correlation between p and variance-mean ratio.

First, define a function that do:

1. Double the cell number
2. Determine how many cells changed their state of EGFR in the cycle of division
3. Repeat the dividing and transformation for *n* times (in this case 12)

```{r define-cell-divide}
divide <- function(a, b, p){
  # Parameters
  # a: initial EGFR+ number
  # b: initial EGFR- number
  # p: the transformation chance from EGFR+ to EGFR-
  ## Given the known relation ship between p and q (q = 19p), we need to check whether q > 1
  ## before we start
  if (p * 19 > 1) {
    stop("Your p should not exceed 0.0526, or q would exceed 1\n(Note p and q are probabilities.)")
  }
  
  # Assign local variables first to make for-loop easier
  this_a <- a
  this_b <- b
  
  for (division in seq(12)) {
    # First the cells divide
    this_a <- 2 * this_a
    this_b <- 2 * this_b
    
  
    # Second the cells decide whether they transform
    # The occurance of transformation should obey the Poisson distribution
    ## Besides that, we need a way to prevent rpois() from drawing more than what we
    ## have (this_a < a_to_b)
    a_to_b <- min(this_a, rpois(n = 1, lambda = this_a * p))
    b_to_a <- min(this_b, rpois(n = 1, lambda = this_b * p * 19))
    
    
    # Count the final number of cells in this loop
    this_a <- this_a - a_to_b + b_to_a
    this_b <- this_b + a_to_b - b_to_a
  }
  return(c(this_a, this_b))
}
```

Then, we run this function 500 times with different value of p.

```{r graded-p-value}
# Note that q = 19 * p
# so p must stay within [0,1/19)
## Generate a numerical vector from 0 to 1/19 stepping 0.001
p_range <- seq(from = 0, to = 1/19, by = 0.001)

# Pre-allocate a matrix to store calculated variance-mean ratio for each p
result <- matrix(ncol = 4, nrow = length(p_range))
colnames(result) <- c("p", "mean_cells", "variance_cells", "variance_mean_ratio")

# Run division() 500 times for each p
for (p in p_range) {
  rep_500 <- sapply(seq(500), function(x) divide(1,0,p))
  mean_pos <- mean(rep_500[1,]) # Count the EGFR+ cells
  var_pos <- var(rep_500[1,]) # Count the EGFR+ cells
  # Save the results and leave space for later calculation of VMR
  result[which(p_range == p), ] <- c(p, mean_pos, var_pos, NA)
}

# Calculating variance over mean
result[,4] <- result[,3]/result[,2]
```
```{r hidden-kable, echo = F}
knitr::kable(head(result), digits = c(3, 2, 2))
```

This would give us some idea about whether a particular p corresponds to a specific variance-mean ratio in this experiment...
```{r load-packages, include=F}
library(ggplot2)
library(cowplot)
```
```{r VMR-plot}
# Turn the result matrix into data frame for ggplot2
result_df <- as.data.frame(result)
vmr_plot <- ggplot(result_df, aes(x = p, y = variance_mean_ratio)) + geom_point() +
  ggtitle("P value versus variance-to-mean ratio") +
  labs(x = "Probability of EGFR+ -> EGFR-",
       y = "Variance-to-Mean Ratio of the Number of\nEGFR+ Cell after Clonal Expansion")
plot(vmr_plot)
```

With the increase of _p_, the variance-to-mean ratio first surges up and then gradually weans down. It is intuitive to imagine when $p\approx0$, very little transformation happens, and thus most of the population stays homogeneous with a low dispersion. On the other hand, when the transformation is sparse, whether transformation happening at the early division would have a much larger impact, and _p_ is not high enough to averge out this effect. Lastly, when _p_ is high, transformation happens offen, and the cells are quickly entering equilibrium where the ratio of the rate constant dominates.

This simulation is actually a general version of [Luria-Delbrück experiment](https://en.wikipedia.org/wiki/Luria–Delbrück_experiment). If we let _q_ be zero, it is the spontaneous mutation condition in this classic experiment.

This iconic variability could also be applied to distinguish whether a phenotype is heritable (a state change with moderately low _p_) or transient (a state change with high _p_). For example, figure 1a in [this article](https://www.nature.com/articles/nature22794) examplified the test.
